Classical Planning Algorithms ¡ on the Atari Video Games

### 2014-11-30 icaps15-paper/paper.tex
added new table. Still need to change the experimental section. This change is towards the workshop camera ready.
Miquel's final iteration

\newcommand{\numgames}{$48$}

\newcommand{\Assault}{\propername{Assault}}
\newcommand{\Asterix}{\propername{Asterix}}
\newcommand{\Asteroids}{\propername{Asteroids}}
\newcommand{\Atlantis}{\propername{Atlantis}}
\newcommand{\BankHeist}{\propername{Bank Heist}}
\newcommand{\BattleZone}{\propername{Battle Zone}}
\newcommand{\BeamRider}{\propername{Beam Rider}}
\newcommand{\Berzerk}{\propername{Berzerk}}
\newcommand{\Bowling}{\propername{Bowling}}
\newcommand{\Boxing}{\propername{Boxing}}
\newcommand{\Breakout}{\propername{Breakout}}
\newcommand{\Carnival}{\propername{Carnival}}
\newcommand{\Centipede}{\propername{Centipede}}
\newcommand{\ChopperCommand}{\propername{Chopper Command}}
\newcommand{\CrazyClimber}{\propername{Crazy Climber}}
\newcommand{\DoubleDunk}{\propername{Double Dunk}}
\newcommand{\Enduro}{\propername{Enduro}}
\newcommand{\FishingDerby}{\propername{Fishing Derby}}
\newcommand{\Freeway}{\propername{Freeway}}
\newcommand{\Frostbite}{\propername{Frostbite}}
\newcommand{\Gravitar}{\propername{Gravitar}}
\newcommand{\Hero}{\propername{Hero}}
\newcommand{\JamesBond}{\propername{James Bond}}
\newcommand{\JourneyEscape}{\propername{Journey Escape}}
\newcommand{\Kangaroo}{\propername{Kangaroo}}
\newcommand{\Krull}{\propername{Krull}}
\newcommand{\KungFuMaster}{\propername{Kung Fu Master}}
\newcommand{\Montezuma}{\propername{Montezuma Revenge}}
\newcommand{\MsPacman}{\propername{Ms Pacman}}
\newcommand{\NameThisGame}{\propername{Name This Game}}
\newcommand{\Pong}{\propername{Pong}}
\newcommand{\Pooyan}{\propername{Pooyan}}
\newcommand{\PrivateEye}{\propername{Private Eye}}
\newcommand{\Qbert}{\propername{Q*Bert}}
\newcommand{\RiverRaid}{\propername{Riverraid}}
\newcommand{\RobotTank}{\propername{Robot Tank}}
\newcommand{\Seaquest}{\propername{Seaquest}}
\newcommand{\SpaceInvaders}{\propername{Space Invaders}}
\newcommand{\StarGunner}{\propername{Star Gunner}}
\newcommand{\Tennis}{\propername{Tennis}}
\newcommand{\TimePilot}{\propername{Time Pilot}}
\newcommand{\Tutankham}{\propername{Tutankham}}
\newcommand{\UpNDown}{\propername{Up And Down}}
\newcommand{\Venture}{\propername{Venture}}
\newcommand{\VideoPinball}{\propername{Video Pinball}}
\newcommand{\WizardOfWor}{\propername{Wizard Of Wor}}
\newcommand{\Zaxxon}{\propername{Zaxxon}}

@@

 \begin{document}
% The file aaai.sty is the style file for AAAI Press
% proceedings, working notes, and technical reports.
%
\pdfinfo{
/Title (Classical Planning Algorithms on the Atari Video Games)
/Author ( )}
% (Nir Lipovetzky, Miquel Ramirez, Hector Geffner)}

@@

\author{\textbf{ID:} 52 \ \ ; \ \ \textbf{Keywords:} classical planning, video games}

% \author{Nir Lipovetzky \\
%       The University of Melbourne \\
%    Melbourne, Australia\\
%    {\normalsize\url{@unimelb.edu.au}}
%    \And
%    Miquel Ramirez \\
%    Australian National University \\
%    Canberra, Australia\\
%   {\normalsize\url{@gmail.com}}
%    \And
%    Hector Geffner \\
%    ICREA \&  Universitat Pompeu Fabra \\
%    Barcelona, SPAIN \\
%    {\normalsize\url{@upf.edu}}\thanks{firstname.lastname}
% }

@@

The Iterated Width  (\IW) algorithm has been introduced as a classical planning algorithm
that takes a planning problem as an input, and computes an action sequence that solves the problem
as the output \cite{nir:ecai2012}. The algorithm however
applies to a broader range of problems. We will characterize such
problems  by means of  a finite and discrete set of states (the state space)
that correspond to vectors of size $n$. Namely, the
states are  \emph{structured} or \emph{factored} , and we take each of the locations in the vector
to represent a variable $X_i$, and the value at that vector location to
represent the value $x_j$ of variable $X_i$ in the state.
In addition to the state space, a problem is defined by an initial state $s_0$,
a set of actions applicable in each state, a
transition function $f$ such that  $s' = f(a,s)$ is
the state that results from applying action $a$  to the state $s$,
and rewards $r(a,s)$ represented by real numbers that result from  applying action $a$ in state $s$.
The transition and reward functions do not need to be known \emph{a priori}, yet in that case, the
state and  reward that results from the application of an action in a state needs to be \emph{observable}.
The task is to compute an action sequence $a_0, \ldots, a_m$ for a large horizon $m$
that generates a state sequence $s_0, \ldots, s_{m+1}$ that  maximizes the accumulated
reward $\sum_{i=0}^{m} r(a_i,s_i)$, or that provides a good approximation.

@@

Formally, the \emph{width} $w(P)$ of a problem $P$ is $i$
iff $i$ is the minimum positive integer for which
there is a sequence of tuples (sets) with at most $i$ atoms each, $t_0, t_1, \ldots, t_n$
such that 1)~$t_0$ is true in the initial state of $P$, 2)~any shortest plan $\pi$  that achieves $t_i$ in $P$
can be extended into a shortest plan that achieves $t_{i+1}$ by extending   $\pi$ with one action, and
3)~any shortest plan that achieves $t_n$ is also a shortest plan for $P$.

One way to understand this definition is that the problem  width $w(P)$ is at most  $i$ if  there is a
``trail of  stepping stones''  $t_0, t_1, \ldots, t_n$ to reach the problem goal such that
A)~these ``stepping stones''  contain at most $i$ atoms each, B)~each  stepping stone $t_{i+1}$ is
at distance 1 from the previous one $t_i$ when $t_i$ has been reached following the ``trail'',
C)~the ``trail''  preserves ``optimality''; i.e., no  tuple $t_i$ can be reached in less than $i$ steps.

While this notion of width and the iterated  width algorithms that are based on it  have been designed
for problems where a goal state needs to be reached, the notions remain
relevant  in  optimization problems as well. Indeed, if a  good path is made of
states $s_i$ each of which has a low width, \IW\ can be made
to find such path in low polynomial time for a small value of the $k$ parameter.
Later on we will discuss  a slight  change  required in \IW\ to enforce this property.
% which is actually  not yet implemented in the code that is tested below.

@@

The number of nodes \emph{generated} by $\IW(1)$ is $n \times D \times b$ in the worst case,
where $n$ is the number of variables, $D$ is the size of their domains, and $b$ is the number of
actions per  state.  This same number in breadth-first search is not linear in $n$
but exponential. For the Atari games,  $n=128$, $D=256$, and $b=18$, so that the product is
equal to $589,824$, which is large but feasible. On the other hand, the number
of nodes generated by $\IW(2)$ in the worst case is  $(n \times D)^2 \times b$, which is
equal to $1,073,741,824$ which is too large, forcing us to consider only a tiny fraction of
such states. For classical planning problems, the growth in the number of nodes from
$\IW(1)$ to $\IW(2)$ is not that large, as the variables are boolean. Indeed, we could have
taken the state vector for the Atari games as a vector of 1024 boolean variables,
and apply these algorithms to that representation. The results would be different.
In such a case, $\IW(1)$ and $\IW(2)$  would generate $n' \times D' \times b$ and
$(n' \times D')^2 \times b$ states, which for $n'=1024$ and $D=2$
represent  $36,864$ and $7,5497,472$ states respectively.
These are feasible numbers, but we haven't used  this representation 
under the assumption that the correlations among the bits in each one of
the 128 words of the state vectors are meaningful.
In summary, from the basic \IW\ algorithm we are testing only the first
call $\IW(1)$.

@@

For  the experiments below, we added two simple variations to  $\IW(1)$ and \BFS.
First, in the breadth-first search underlying $\IW(1)$, we generate the children in
random order. This makes the executions that results from the
$\IW(1)$ lookahead less susceptible to fall into loops; a potential
problem in local search algorithms with no memory or learning.
Second, a discount factor $\gamma=0.995$ is used in both algorithms
for discounting future rewards like in \UCT.
For this, each state $s$ keeps its depth $d(s)$ in the search tree,
and if state s' is the child of state $s$ and action $a$,
$R(s')$ is set to $R(s) + \gamma^{d(s)+1} r(a,s)$.
The  discount factor  results in a slight
preference for rewards that can be reached earlier,
which is a  reasonable heuristic in  on-line settings
based on lookahead searches.

@@

\input{games-table}
We tested \IW(1) and \BFS\ over \numgames\
 of the $55$ games considered in
\cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
Entries marked as  ``--'' in the table are   still running at the time of submission. 
Each run can take up to $24$ hours of CPU time and there are $10$ runs per game and algorithm. 
Eventually all of  the  $55$ games will  be included in the evaluation.}
Table~\ref{performance} shows the performance of $\IW(1)$ and \BFS\
% , characterized by the average score obtained over all episodes, the standard deviation
% of the score obtained and time (in seconds) per lookahead,
in comparison  with breadth-first search (\BRFS) and \UCT.
Videos of selected games played by $\IW(1)$, \BFS, and \UCT\ can be 
seen in Youtube.\footnote{\url{http://www.youtube.com/playlist?list=PLXpQcXUQ_CwenUazUivhXyYvjuS6KQOI0}.}
The discount factor used by  all the algorithms is $\gamma=0.995$.
The scores reported for \BRFS\ and \UCT\ are taken from BNVB.  
Our experimental setup  follows theirs except that a maximum budget of  $150,000$ simulated frames 
is applied to $\IW(1)$, \BFS, and \UCT. \UCT\ uses this budget by running
 $500$   rollouts of depth $300$. % \footnote{In \UCT,  $500$ base policy rollouts
% of depth $300$ is equivalent to a budget of $150,000$ simulated frames.}.
The bound on the number of simulated frames is like a bound on
lookahead time, as most of the time in the lookahead  is spent
in calls to the emulator for computing the next RAM state. 
This is why the average time per action is similar to all the algorithms
except $\IW(1)$, that due to its aggressive pruning does not always use
the full budget and takes less time per action on average.

Also,  as reported by  BNVB, all of the algorithms 
reuse the sub-tree from the previous lookahead that is rooted in the selected 
child, deleting its siblings and their descendants. More precisely, no calls 
to the emulator are done for transitions that are cached in that sub-tree, 
and such reused frames are not discounted from the budget that is thus a bound 
on the number of ``new'' frames per lookahead. In addition, in $\IW(1)$, the states that 
are ``reused'' from the previous searches are ignored in the computation of the novelty of 
new states so that more states can escape pruning. Otherwise, $\IW(1)$ often
uses a fraction of the  budget.  This is not needed in \BFS\ which does 
no pruning.  $\IW(1)$ and \BFS\ are  limited to  search up to a depth of
$m=1,500$ frames and up to $m=150,000$ frames per root branch. This is
to avoid the search from  going too deep or being too committed to a
single root action. 
% Similarly UCT defines a limit of $m=300$ frames for each rollout. 
% Note that UCT is more likely to have a performance
% loss if the depth is increased, as the number of rollouts would
% decrease in order to maintain the same budget of $150,000$ simulated
% frames.

@@

Table~\ref{performance} shows that both $\IW(1)$ and \BFS\ generally
 outperform \BRFS, which rarely collects any reward in many domains, as 
 the depth of the \BRFS\ search tree results in a lookahead of $0.3$
seconds ($20$ frames or $4$ nodes deep). The notable exception to this
is \Centipede\ where abundant reward can be collected with a shallow
lookahead. On the other hand, both $\IW(1)$ and \BFS\ reach states that 
are up to $350$--$1,500$ frames deep ($70$--$260$ nodes or $6$--$22$ seconds), 
even if $\IW(1)$ does not always use all the simulation frames allocated due to its
pruning. This behaviour can be observed in games such as
\Breakout, \CrazyClimber, \Kangaroo, and \Pooyan, where the average
CPU time for each lookahead is up to $10$ times faster than \BFS. 
Computation time for \UCT\ and \BRFS\ are  similar to \BFS, as the
most expensive part of the computation is the  generation of 
frames through the simulator,  and these three  algorithms 
always use  the full budget.

More interestingly,  \IW(1) outscores \UCT\  in $27$ of the \numgames\ games,
while \BFS\ outscores \UCT\ in $21$.
%\footnote{These numbers can actually
%improve further as 6  games for \IW(1) and 4 games for \BFS\ are still running.}
On the other hand, \UCT\ does better than $\IW(1)$ and \BFS\ in $15$ and $17$ games respectively.
The relative performance between $\IW(1)$ and \BFS\ makes $\IW(1)$  the
best of the two in $22$ games, and  \BFS\  in $12$. In terms of the number of games 
where an algorithm is the overall winner, $\IW(1)$  domainates the others: 
$\IW(1)$ is the overall best in $23$ games, \BFS\ in $11$ games, and \UCT\ in $18$ games. 
Also, \BRFS\ is best in $2$ games (\Centipede, tied up in \Boxing), while the other three algorithms
are tied in another 2 games (\Pong, \Boxing), and $\IW(1)$ and \BFS\ are tied in \TimePilot.

@@

Likewise, in \Freeway\ and \Berzerk\ both $\IW(1)$ and \BFS\ attain a better
score than the baseline semi-random algorithm \emph{Perturb} in
\cite{bellemare:jair2013}, that beats \UCT\ on those games.
%% to say the 2 lines below must add context: .. eg
%%% ``which in *these games* outperforms both \UCT\ and the reinforcement learning methods.''
%% add it if that's the case ..
\emph{Perturb} is a simple algorithm
that selects  a fixed  action with probability $0.95$, and a random
action with probability $0.05$.  For \emph{Perturb}, BNVB do not report 
the average score  but the  best score.  \emph{Perturb} manages to do well in domains where rewards
are deep but can be reached by  repeating the same action.
This is the case of \Freeway, where a chicken has to run to the top of the screen across a ten lane
highway filled with traffic. Every time the chicken gets across (starting at the bottom), there is one
unit of reward. If the chicken is hit by a car, it goes back some lanes. 
In \Freeway, only $12$ out of the $18$ possible actions have an effect:
$6$ actions move the chicken up (up-right, up-left, up-fire, up-right-fire, up-left-fire), 
$6$ actions move the chicken down (down-right, down-left, down-fire, down-right-fire, down-left-fire), 
and $6$ actions do nothing.  \emph{Perturb}  does well in this domain 
when the selected fixed action moves the chicken up.  As noted 
in Table~\ref{performance} and seen in the provided video, \UCT\ does not manage to take the chicken
across the highway at all. 
%
The reason that \UCT\ does not collect any reward is that it needs to move the chicken up
at least 240 times~\footnote{One needs to move the chicken up for at least $4$ seconds ($240$ frames) in order to get across the 
highway.} something that is very unlikely in a random exploration. $\IW(1)$ does not have this limitation and is best in \Freeway.
%% The reason for \UCT\ to perform erratically 
%% is that the roll-outs do not manage to collect any reward. A sequence of actions 
%% with at least $48$ actions embedded that move the chicken up are required in order 
%% to collect a reward. Since the probability for the roll-out to choose an action that moves the chicken up 
%% is $1/3$, with the parameters chosen by BNVB, the likelihood of \UCT\ generating such a sequence of actions 
%% that \emph{also} avoids incoming traffic in a timely fashion is rather low.
%% %% the previous number was not correct; no-op actions do not count, and thus needs
%% %%% 200 more move up actions than move down actions, although this is not accurate either
%% %%%% as move down actions that take chicken below the screen count as no-ops too ..
%% $\IW(1)$ does not have this limitation and is best in \Freeway.

@@

$\IW(1)$ also outperforms the best learning algorithm in
\cite{deep-mind-atari} in the $7$ games considered there, and
\BFS\ does so in $6$ of the $7$ games. Comparing with the scores reported for the reinforcement learning algorithms by BNVB,
we note that both $IW(1)$ and \BFS\ do much better than the best learning algorithm in
those games where the learning algorithms outperform \UCT\ like \Montezuma, \Venture\
and \Bowling.  In \Montezuma\ rewards are very
sparse, deep, and most of the actions lead to losing a life with
no immediate penalty or consequence.  All algorithms achieve $0$ score, 
except for \BFS\ that achieves an average score of $80$, and a score
of $400$ in one of the runs. This means that even \BFS\  is not able to 
consistently find rewards in this game. This game and several other games 
like \Breakout, and \SpaceInvaders\ could be much simpler by adding
negative rewards for losing a life. We have indeed observed that
our planning algorithms do not care much about losing lives
until there is just one life left, when their play noticeably improves.
This can be seen in the videos mentioned above, and suggest a simple form of learning that would be useful to both planners
and reinforcement learning algorithms.

@@

Regarding the Atari games, we identify two problems in all the planning methods
discussed. First, by using the RAM state of the Atari console one can consider
that they are ``cheating'', as they use information that is not available to the human player. And second, they do not get better with experience. In the future, we would like to explore 
whether variations of these algorithms  can be used to play from the information conveyed by the  
 state of screen pixels as opposed to the RAM state, and the potential uses of the new ideas for learning to improve performance.

>>>>

\newcommand{\numgames}{$54$}

\newcommand{\Alien}{\propername{Alien}}
\newcommand{\Amidar}{\propername{Amidar}}
\newcommand{\Assault}{\propername{Assault}}
\newcommand{\Asterix}{\propername{Asterix}}
\newcommand{\Asteroids}{\propername{Asteroids}}
\newcommand{\Atlantis}{\propername{Atlantis}}
\newcommand{\BankHeist}{\propername{Bank Heist}}
\newcommand{\BattleZone}{\propername{Battle Zone}}
\newcommand{\BeamRider}{\propername{Beam Rider}}
\newcommand{\Berzerk}{\propername{Berzerk}}
\newcommand{\Bowling}{\propername{Bowling}}
\newcommand{\Boxing}{\propername{Boxing}}
\newcommand{\Breakout}{\propername{Breakout}}
\newcommand{\Carnival}{\propername{Carnival}}
\newcommand{\Centipede}{\propername{Centipede}}
\newcommand{\ChopperCommand}{\propername{Chopper Command}}
\newcommand{\CrazyClimber}{\propername{Crazy Climber}}
\newcommand{\DemonAttack}{\propername{Demon Attack}}
\newcommand{\DoubleDunk}{\propername{Double Dunk}}
\newcommand{\ElevatorAction}{\propername{Elevator Action}}
\newcommand{\Enduro}{\propername{Enduro}}
\newcommand{\FishingDerby}{\propername{Fishing Derby}}
\newcommand{\Freeway}{\propername{Freeway}}
\newcommand{\Frostbite}{\propername{Frostbite}}
\newcommand{\Gopher}{\propername{Gopher}}
\newcommand{\Gravitar}{\propername{Gravitar}}
\newcommand{\Hero}{\propername{Hero}}
\newcommand{\IceHockey}{\propername{Ice Hockey}}
\newcommand{\JamesBond}{\propername{James Bond}}
\newcommand{\JourneyEscape}{\propername{Journey Escape}}
\newcommand{\Kangaroo}{\propername{Kangaroo}}
\newcommand{\Krull}{\propername{Krull}}
\newcommand{\KungFuMaster}{\propername{Kung Fu Master}}
\newcommand{\Montezuma}{\propername{Montezuma Revenge}}
\newcommand{\MsPacman}{\propername{Ms Pacman}}
\newcommand{\NameThisGame}{\propername{Name This Game}}
\newcommand{\Pong}{\propername{Pong}}
\newcommand{\Pooyan}{\propername{Pooyan}}
\newcommand{\PrivateEye}{\propername{Private Eye}}
\newcommand{\Qbert}{\propername{Q*Bert}}
\newcommand{\RiverRaid}{\propername{Riverraid}}
\newcommand{\RoadRunner}{\propername{Road Runner}}
\newcommand{\RobotTank}{\propername{Robot Tank}}
\newcommand{\Seaquest}{\propername{Seaquest}}
\newcommand{\Skiing}{\propername{Skiing}}
\newcommand{\SpaceInvaders}{\propername{Space Invaders}}
\newcommand{\StarGunner}{\propername{Star Gunner}}
\newcommand{\Tennis}{\propername{Tennis}}
\newcommand{\TimePilot}{\propername{Time Pilot}}
\newcommand{\Tutankham}{\propername{Tutankham}}
\newcommand{\UpNDown}{\propername{Up And Down}}
\newcommand{\Venture}{\propername{Venture}}
\newcommand{\VideoPinball}{\propername{Video Pinball}}
\newcommand{\WizardOfWor}{\propername{Wizard Of Wor}}
\newcommand{\Zaxxon}{\propername{Zaxxon}}

@@

 \begin{document}
% The file aaai.sty is the style file for AAAI Press
% proceedings, working notes, and technical reports.
%
\pdfinfo{
/Title (Classical Planning Algorithms on the Atari Video Games)
/Author (Nir Lipovetzky, Miquel Ramirez, Hector Geffner)}

@@

% \author{\textbf{ID:} 52 \ \ ; \ \ \textbf{Keywords:} classical planning, video games}

 \author{Nir Lipovetzky \\
       The University of Melbourne \\
    Melbourne, Australia\\
%     {\normalsize\url{nir.lipovetzky@unimelb.edu.au}}
%%% footnote doesn't look nice ; this looks better
%% although can't use Nir's unimelb address that is too long :-)
%% 
    {\small\url{nirlipo@gmail.com}}
    \And
    Miquel Ramirez \\
    Australian National University \\
    Canberra, Australia\\
   {\small\url{miquel.ramirez@gmail.com}}
    \And
    Hector Geffner \\
    ICREA \&  Universitat Pompeu Fabra \\
    Barcelona, SPAIN \\
    {\small\url{hector.geffner@upf.edu}}
 %%%%% \thanks{firstname.lastname}
}

@@

The Iterated Width  (\IW) algorithm has been introduced as a classical planning algorithm
that takes a planning problem as an input, and computes an action sequence that solves the problem
as the output \cite{nir:ecai2012}. The algorithm however
applies to a broader range of problems. We will characterize such
problems  by means of  a finite and discrete set of states (the state space)
that correspond to vectors of size $n$. Namely, the
states are  \emph{structured} or \emph{factored} , and we take each of the locations in the vector
to represent a variable $X_i$, and the value at that vector location to
represent the value $x_j$ of variable $X_i$ in the state.
In addition to the state space, a problem is defined by an initial state $s_0$,
a set of actions applicable in each state, a
transition function $f$ such that  $s' = f(a,s)$ is
the state that results from applying action $a$  to the state $s$,
and rewards $r(a,s)$ represented by real numbers that result from  applying action $a$ in state $s$.
The transition and reward functions do not need to be known \emph{a priori}, yet in that case, the
state and  reward that results from the application of an action in a state need to be \emph{observable}.
The task is to compute an action sequence $a_0, \ldots, a_m$ for a large horizon $m$
that generates a state sequence $s_0, \ldots, s_{m+1}$ that  maximizes the accumulated
reward $\sum_{i=0}^{m} r(a_i,s_i)$, or that provides a good approximation.

@@

Formally, the \emph{width} $w(P)$ of a problem $P$ is $i$ iff $i$ is the minimum positive integer for which
there is a sequence $t_0, t_1, \ldots, t_n$ of  atom sets  $t_k$  with at most $i$ atoms each, 
such that 1)~$t_0$ is true in the initial state of $P$, 2)~any shortest plan $\pi$  that achieves $t_k$ in $P$
can be extended into a shortest plan that achieves $t_{k+1}$ by extending   $\pi$ with one action, and
3)~any shortest plan that achieves $t_n$ is  a shortest plan for achieving the goal of $P$.

One way to understand this definition is that the problem  width $w(P)$ is at most  $i$ if  there is a
``trail of  stepping stones''  $t_0, t_1, \ldots, t_n$ to reach the problem goal such that
A)~these ``stepping stones''  contain at most $i$ atoms each, B)~each  stepping stone $t_{k+1}$ is
at distance 1 from the previous one $t_k$ when $t_k$ has been reached following the ``trail'',
C)~the ``trail''  preserves ``optimality''; i.e., no  tuple $t_k$ can be reached in less than $k$ steps.

While this notion of width and the iterated  width algorithms that are based on it  have been designed
for problems where a goal state needs to be reached, the notions remain
relevant  in  optimisation problems as well. Indeed, if a  good path is made of
states $s_i$ each of which has a low width, \IW\ can be made
to find such path in low polynomial time for a small value of the $k$ parameter.
Later on we will discuss  a slight  change  required in \IW\ to enforce this property.
% which is actually  not yet implemented in the code that is tested below.

@@

The number of nodes \emph{generated} by $\IW(1)$ is $n \times D \times b$ in the worst case,
where $n$ is the number of problem variables, $D$ is the size of their domains, and $b$ is the number of
actions per  state.  This same number in a  breadth-first search is not linear in $n$
but exponential. For the Atari games,  $n=128$, $D=256$, and $b=18$, so that the product is
equal to $589,824$, which is large but feasible. On the other hand, the number
of nodes generated by $\IW(2)$ in the worst case is  $(n \times D)^2 \times b$, which is
equal to $1,073,741,824$ which is too large, forcing us to consider only a tiny fraction of
such states. For classical planning problems, the growth in the number of nodes from
$\IW(1)$ to $\IW(2)$ is not that large, as the variables are boolean. Indeed, we could have
taken the state vector for the Atari games as a vector of 1024 boolean variables,
and apply these algorithms to that representation. The results would be different.
In such a case, $\IW(1)$ and $\IW(2)$  would generate $n' \times D' \times b$ and
$(n' \times D')^2 \times b$ states, which for $n'=1024$ and $D=2$
represent  $36,864$ and $7,5497,472$ states respectively.
These are feasible numbers, but we haven't used  this representation 
under the assumption that the correlations among the bits in each one of
the 128 words of the state vectors are meaningful.
In summary, from the basic \IW\ algorithm we are testing only the first
call $\IW(1)$.

@@

For  the experiments below, we added two simple variations to  $\IW(1)$ and \BFS.
First, in the breadth-first search underlying $\IW(1)$, we generate the children in
random order. This makes the executions that result from the
$\IW(1)$ lookahead less susceptible to be trapped into loops; a potential
problem in local search algorithms with no memory or learning.
Second, a discount factor $\gamma=0.995$ is used in both algorithms
for discounting future rewards like in \UCT.
For this, each state $s$ keeps its depth $d(s)$ in the search tree,
and if state s' is the child of state $s$ and action $a$,
$R(s')$ is set to $R(s) + \gamma^{d(s)+1} r(a,s)$.
The  discount factor  results in a slight
preference for rewards that can be reached earlier,
which is a  reasonable heuristic in  on-line settings
based on lookahead searches.

@@

\input{games-table}
We tested \IW(1) and \BFS\ over \numgames\  of the $55$ games 
considered in \cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
We left out  \textsc{Skiing} as the reported figures apparently use a  different reward structure.}
%% check
% \footnote{
% Entries marked as  ``--'' in the table are   still running at the time of submission. 
% Each run can take up to $24$ hours of CPU time and there are $10$ runs per game and algorithm. 
% Eventually all of  the  $55$ games will  be included in the evaluation.}
Table~\ref{performance} shows the performance of $\IW(1)$ and \BFS\
in comparison  with breadth-first search (\BRFS) and \UCT.
Videos of selected games played by $\IW(1)$, \BFS, and \UCT\ can be 
seen in Youtube.\footnote{\url{http://www.youtube.com/playlist?list=PLXpQcXUQ_CwenUazUivhXyYvjuS6KQOI0}.}
The discount factor used by  all the algorithms is $\gamma=0.995$.
The scores reported for \BRFS\ and \UCT\ are taken from BNVB.  
Our experimental setup  follows theirs except that a maximum budget of  $150,000$ simulated frames 
is applied to $\IW(1)$, \BFS, and \UCT. \UCT\ uses this budget by running
 $500$   rollouts of depth $300$. % \footnote{In \UCT,  $500$ base policy rollouts
% of depth $300$ is equivalent to a budget of $150,000$ simulated frames.}.
The bound on the number of simulated frames is like a bound on
lookahead time, as most of the time in the lookahead  is spent
in calls to the emulator for computing the next RAM state. 
This is why the average time per action is similar to all the algorithms
except $\IW(1)$, that due to its aggressive pruning does not always use
the full budget and takes less time per action on average.

Also,  as reported by  BNVB, all of the algorithms 
reuse the frames in the sub-tree of the previous lookahead that is rooted in the selected 
child, deleting its siblings and their descendants. More precisely, no calls 
to the emulator are done for transitions that are cached in that sub-tree, 
and such reused frames are not discounted from the budget that is thus a bound 
on the number of \emph{new}  frames per lookahead. In addition, in $\IW(1)$, the states that 
are reused  from the previous searches are ignored in the computation of the novelty of 
new states so that more states can escape pruning. Otherwise, $\IW(1)$ often
uses a fraction of the  budget.  This is not needed in \BFS\ which does 
no pruning.  $\IW(1)$ and \BFS\ are  limited to  search up to a depth of
$m=1,500$ frames and up to $m=150,000$ frames per root branch. This is
to avoid the search from  going too deep or being too committed to a
single root action. 
% Similarly UCT defines a limit of $m=300$ frames for each rollout. 
% Note that UCT is more likely to have a performance
% loss if the depth is increased, as the number of rollouts would
% decrease in order to maintain the same budget of $150,000$ simulated
% frames.

@@

Table~\ref{performance} shows that both $\IW(1)$ and \BFS\ outperform \BRFS, which  rarely collects
reward in  many domains as  the depth of the \BRFS\ search tree results in a lookahead of $0.3$
seconds ($20$ frames or $4$ nodes deep). The notable exception to this
is \Centipede\ where abundant reward can be collected with a shallow
lookahead. On the other hand, both $\IW(1)$ and \BFS\ normally 
reach states that are up to $350$--$1500$ frames deep ($70$--$260$ nodes or 
$6$--$22$ seconds),  even if $\IW(1)$ does not always use all the simulation frames allocated due to its
agressive pruning. This can be observed in games such as \Breakout, \CrazyClimber, \Kangaroo, and \Pooyan, where the average
CPU time for each lookahead is up to $10$ times faster than \BFS.  Computation time for \UCT\ and \BRFS\ are  similar to \BFS, as the
most expensive part of the computation is the  generation of frames through the simulator,  and these three  algorithms 
always use  the full budget.

More interestingly,  \IW(1) outscores \UCT\  in $31$ of the \numgames\ games,
while \BFS\ outscores \UCT\ in $26$.
%\footnote{These numbers can actually
%improve further as 6  games for \IW(1) and 4 games for \BFS\ are still running.}
On the other hand, \UCT\ does better than $\IW(1)$ and \BFS\ in $19$ and $25$ games respectively.
The relative performance between $\IW(1)$ and \BFS\ makes $\IW(1)$  the
best of the two in $34$ games, and  \BFS\  in $16$. In terms of the number of games 
where an algorithm is the best, % $\IW(1)$  domainates the others: 
$\IW(1)$ is  the best in $26$ games, \BFS\ in $13$ games, and \UCT\ in $19$ games. 
Also, \BRFS\ is best in $2$ games (\Centipede, tied up in \Boxing), while the other three algorithms
are tied in another 2 games (\Pong, \Boxing). 
% , and $\IW(1)$ and \BFS\ are tied in \TimePilot.

@@

Likewise, in \Freeway\ and \Berzerk\ both $\IW(1)$ and \BFS\ attain a better
score than the baseline semi-random algorithm \emph{Perturb} in
\cite{bellemare:jair2013}, that beats \UCT\ on those games.
%% to say the 2 lines below must add context: .. eg
%%% ``which in *these games* outperforms both \UCT\ and the reinforcement learning methods.''
%% add it if that's the case ..
\emph{Perturb} is a simple algorithm
that selects  a fixed  action with probability $0.95$, and a random
action with probability $0.05$.  For \emph{Perturb}, BNVB do not report 
the average score  but the  best score.  \emph{Perturb} manages to do well in domains where rewards
are deep but can be reached by  repeating the same action.
This is the case of \Freeway, where a chicken has to run to the top of the screen across a ten lane
highway filled with traffic. Every time the chicken gets across (starting at the bottom), there is one
unit of reward. If the chicken is hit by a car, it goes back some lanes. 
In \Freeway, only $12$ out of the $18$ possible actions have an effect:
$6$ actions move the chicken up (up-right, up-left, up-fire, up-right-fire, up-left-fire), 
$6$ actions move the chicken down (down-right, down-left, down-fire, down-right-fire, down-left-fire), 
and $6$ actions do nothing.  \emph{Perturb}  does well in this domain 
when the selected fixed action moves the chicken up.  As noted 
in Table~\ref{performance} and seen in the provided video, \UCT\ does not manage to take the chicken
across the highway at all. 
%
The reason that \UCT\ does not collect any reward is that it needs to move the chicken up
at least 240 times\footnote{One needs to move the chicken up for at least $4$ seconds ($240$ frames) in order to get across the 
highway.} something that is very unlikely in a random exploration. $\IW(1)$ does not have this limitation and is best in \Freeway.
%% The reason for \UCT\ to perform erratically 
%% is that the roll-outs do not manage to collect any reward. A sequence of actions 
%% with at least $48$ actions embedded that move the chicken up are required in order 
%% to collect a reward. Since the probability for the roll-out to choose an action that moves the chicken up 
%% is $1/3$, with the parameters chosen by BNVB, the likelihood of \UCT\ generating such a sequence of actions 
%% that \emph{also} avoids incoming traffic in a timely fashion is rather low.
%% %% the previous number was not correct; no-op actions do not count, and thus needs
%% %%% 200 more move up actions than move down actions, although this is not accurate either
%% %%%% as move down actions that take chicken below the screen count as no-ops too ..
%% $\IW(1)$ does not have this limitation and is best in \Freeway.

@@

$\IW(1)$ also outperforms the best learning algorithm in
\cite{deep-mind-atari} in the $7$ games considered there, and
\BFS\ does so in $6$ of the $7$ games. Comparing with the scores reported for the reinforcement learning algorithms by BNVB,
we note that both $\IW(1)$ and \BFS\ do much better than the best learning algorithm in
those games where the learning algorithms outperform \UCT\ like \Montezuma, \Venture\
and \Bowling.  In \Montezuma\ rewards are very
sparse, deep, and most of the actions lead to losing a life with
no immediate penalty or consequence.  All algorithms achieve $0$ score, 
except for \BFS\ that achieves an average score of $540$, and a score
of $2500$ in one of the runs. This means however that even \BFS\  is not able to 
consistently find rewards in this game. This game and several others
like \Breakout\ and \SpaceInvaders\ could be much simpler by adding
negative rewards for losing a life. We have indeed observed that
our planning algorithms do not care much about losing lives
until there is just one life left, when their play noticeably improves.
This can be seen in the videos mentioned above, and suggest a simple form of learning that would be useful to both planners
and reinforcement learning algorithms.

@@

Concerning the Atari games, there are two problems in all the planning methods
discussed. First, by using the RAM state of the Atari console one can consider
that they are ``cheating'', as they use information that is not available to the human player. And second, they do not get 
better with experience. In the future, we would like to explore 
whether variations of these algorithms  can be used to play from the information conveyed by the  
state of the screen pixels as opposed to the RAM state, and the potential uses of the new ideas for learning,
such the use of the  novelty measures for shaping rewards.

\medskip

\noindent \textbf{Acknowledgments.} We thank the people involved in the creation and maintenance of the 
Arcade Learning Environment  that have made this work possible.
